---
title: "Data Science Capstone"
author: "jbratanov"
date: "Monday, August 22, 2016"
output: html_document
---
# Synopsis
We will be doing Exploratory data analysis of data from SwiftKey, a company which is known for their application on cell phones that learn your word patterns as you type messages.  Our goal is to build a similar predictive text pattern algorithm as the Swiftkey solution using Natural Language Processing to predict the next word or set of words as users type real time.

# Data Processing

### Dataset Information
The data is from a corpus called HC Corpora and is located at www.corpora.heliohost.org.  The readme file for the dataset used is located at www.corpora.heliohost.org/aboutcorpus.html.

### Dataset Download Information
In order to speed up retrieval of the training data to be used by the Capstone project, I manually downloaded the training data located at https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip.

The en_us based datasets were unzipped into the capstone project data directory as: 

+ en_US.news.txt
+ en_US.twitter.txt
+ en_US.blogs.txt


```{r cacheDataRead, cache=TRUE}
# Load files using cache option, as re-reading during test would take too long
# Included skilNul=TRUE statement to get rid of nulls from first pass.
newsText <- readLines("../data/en_US.news.txt", skipNul = TRUE)
twitterText <- readLines("../data/en_US.twitter.txt", skipNul = TRUE)
blogText <- readLines("../data/en_US.blogs.txt", skipNul = TRUE)
```
### Dataset Analysis
```{r cacheDataRead2, cache=TRUE}
library(stringr)

# Show first few rows of data
head(blogText)
head(newsText)
head(twitterText)

# Get number of lines
blogLen <- length(blogText)
newsLen <- length(newsText)
twitLen <- length(twitterText)
paste("Number of lines:", "Blogger Text=", blogLen, "News Text=", newsLen, "Twitter Text=", twitLen)

# Get number of words
blogWords <- sum(sapply(strsplit(blogText, "\\s+"), length))
newsWords <- sum(sapply(strsplit(newsText, "\\s+"), length))
twitWords <- sum(sapply(strsplit(twitterText, "\\s+"), length))

# Get number of characters
blogSum <- sum(nchar(blogText))
newsSum <- sum(nchar(newsText))
twitSum <- sum(nchar(twitterText))
paste("Number of chars:", "Blogger Text=", blogSum, "News Text=", newsSum, "Twitter Text=", twitSum)

# Get average line size
paste("Average line size in characters:", "Blogger Text=", as.integer(blogSum/blogLen),
      "News Text=", as.integer(newsSum/newsLen),
      "Twitter Text=", as.integer(twitSum/twitLen))

# Create DF to store info
name <- c('Blog', 'News', 'Twitter')
lines <- c(blogLen, newsLen, twitLen)
words <- c(blogWords, newsWords, twitWords)
chars <- c(blogSum, newsSum, twitSum)
avgLineSize <- c(as.integer(blogSum/blogLen), as.integer(newsSum/newsLen), as.integer(twitSum/twitLen))
textInfo <- data.frame(name, lines, words, chars, avgLineSize)

# Plot data

# Create a summary grid of data
library(gridExtra)
grid.table(textInfo)

# Number of Lines Plot
par(cex=.75)
plotLines <- barplot(textInfo$lines/1000, main="Number of Lines in (Thousands)",
                     horiz=TRUE, names.arg=textInfo$name, col=c("yellow", "red", "green"),
                     ylab="Corpus Source", cex.lab=1, cex.main=2, xlim=c(0, 2500))
text(textInfo$lines/1000+50, 1:3, as.integer(textInfo$lines/1000), cex=1)

# Number of Words Plot
par(cex=.75)
plotLines <- barplot(textInfo$words/1000000, main="Number of Words in (Millions)",
                     horiz=TRUE, names.arg=textInfo$name, col=c("yellow", "red", "green"),
                     ylab="Corpus Source", cex.lab=1, cex.main=2, xlim=c(0, 40))
text(textInfo$words/1000000+1, 1:3, as.integer(textInfo$words/1000000), cex=1)

# Number of characters Plot
par(cex=.75)
plotLines <- barplot(textInfo$chars/1000000, main="Number of Characters in (Millions)",
                     horiz=TRUE, names.arg=textInfo$name, col=c("yellow", "red", "green"),
                     ylab="Corpus Source", cex.lab=1, cex.main=2, xlim=c(0, 300))
text(textInfo$chars/1000000+5, 1:3, as.integer(textInfo$chars/1000000), cex=1)

# Average line size Plot
par(cex=.75)
plotLines <- barplot(as.integer(textInfo$avgLineSize), main="Average Line Size",
                     horiz=TRUE, names.arg=textInfo$name, col=c("yellow", "red", "green"),
                     ylab="Corpus Source", cex.lab=1, cex.main=2, xlim=c(0, 300))
text(as.integer(textInfo$avgLineSize)+5, 1:3, as.integer(textInfo$avgLineSize), cex=1)



```



